Cheminformatics, the Computer Science of Chemical Discovery, Turning Open-Source

### 2012-03-18 resubmit2/cheminformatics_CACM_revision.tex
Put in the start of the third and final revision

\begin{figure}
\centering
\includegraphics[height=2in]{chemical-structures.png}
\caption{A hierarchy of chemical structure representations, increasing
  in information content from left to right and top to bottom. A
  chemist's traditional 2D depiction (top left); a labeled graph (top
  right -- `A' corresponds to aromatic atoms, `O' to oxygens and `On' to
  oxygens or nitrogens, unlabeled nodes have an implicit label of `C'; a 3D
  conformation (bottom left) and a surface representation (bottom right). All show differing,
  valid aspects of the same molecule, and are suited for
  different purposes, with representation choice guided by
  intended use.}
\label{figure:chemical-structures}
\end{figure}

@@

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in isolation.
Traditionally, predictive models have focused on a single structure
for a small molecule and ignored the receptor. Yet, small molecules
can exist in multiple tautomeric forms conformations. 
Enhancing the accuracy of predictions
will ideally require that the 3D geometries of that molecule be taken
into account and the receptor be considered as far as possible.  
While some relevant low-energy conformers of small
molecules may be accessible in crystallographic databases, this is
not always the case. Though it is now possible to generate reasonable
low energy conformations \emph{ab initio}, the
``biologically relevant'' conformation might differ from the lowest
energy conformation of the molecule considered in isolation,
necessitating the need for conformational search.  Multi-conformer
modeling has been addressed in the 4D-QSAR methodology described by
Hopfinger \textit{et al.}  \cite{Albuquerque:1998ys}.  Recent
techniques such as \emph{multiple-instance learning} could also be
applied to the multi-conformer problem.

>>>>

\begin{figure}
\centering
\includegraphics[height=2in]{chemical-structures.png}
\caption{A hierarchy of chemical structure representations, increasing
  in information content from left to right and top to bottom. A
  chemist's traditional 2D depiction (top left); a labeled graph (top
  right -- `A' corresponds to aromatic atoms, `O' to oxygens and `On' to
  oxygens or nitrogens, unlabeled nodes have an implicit label of `C'); a 3D
  conformation (bottom left) and a surface representation (bottom right). All show differing,
  valid aspects of the same molecule, and are suited for
  different purposes, with representation choice guided by
  intended use.}
\label{figure:chemical-structures}
\end{figure}

@@

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in isolation.
Traditionally, predictive models have focused on a single structure
for a small molecule and ignored the receptor. Yet, small molecules
can exist in multiple tautomeric forms and conformations. 
Enhancing the accuracy of predictions
will ideally require that the 3D geometries of that molecule be taken
into account and the receptor be considered as far as possible.  
While some relevant low-energy conformers of small
molecules may be accessible in crystallographic databases, this is
not always the case. Though it is now possible to generate reasonable
low energy conformations \emph{ab initio}, the
``biologically relevant'' conformation might differ from the lowest
energy conformation of the molecule considered in isolation,
necessitating the need for conformational search.  Multi-conformer
modeling has been addressed in the 4D-QSAR methodology described by
Hopfinger \textit{et al.}  \cite{Albuquerque:1998ys}.  Recent
techniques such as \emph{multiple-instance learning} could also be
applied to the multi-conformer problem.

### 2012-03-18 resubmit2/cheminformatics_second_CACM_revision.tex
Put in the start of the third and final revision

\documentclass{../sig-alternate}

@@

@@

So why is chemical information important and why should we care about
its public availability -- and how does this relate to the field of
computer science? Cheminformatics is widely used in fields 
from agrochemical research to the design of
novel materials -- but we have chosen to use drug discovery as the context
of this article due to its central relevance for human well-being.
The art and science of drug discovery centers around small
molecules, with a core component of methods involving 
techniques and algorithms to handle, analyse and interpret
chemical structure information. 
Unfortunately, drug discovery research is slow, expensive and prone to failure. 
Public availability of chemical information and
tools to handle it are important, as the more information available to each
researcher, the better the chances of avoiding the many causes of attrition. 
Computer science research is highly relevant to managing the \textit{volume} 
and \textit{complexity} of chemical information. 
A single database such as PubChem contains more than 34 million
chemical structure records), along with an even larger number of
\emph{annotations}, such as synonyms (different names for the same molecule; 
PubChem stores 50.1 million synonym for 19.6 million compounds),
known targets (drugs with which the molecule is known to interact),
mode of action (how the molecule interacts with its targets),
regulatory approval status and much more.

@@

Chemical and bioactivity
databases such as ChEMBL and PubChem are freely
available and contain in the order of millions (ChEMBL) and tens of
millions (PubChem) of data points. Integration of these disparate
data is essential for researchers to obtain the fullest possible
perspective on what is presently known, tracking ongoing advances in
science as these become available. Integration of data
across chemical databases is a challenge due to sheer data volume and
due to the difficulties in normalization of chemical and bioactivity
data. Molecular graphs must be captured
in a machine-readable fashion. It is also necessary to be able to 
search in these data, for example multi-labeled graphs, where graph labels can change
from database to database. Not only can labels differ between databases, 
but a chemical graph can be encoded in multiple ways, depending on how 
the nodes are ordered, resulting in multiple representations of the same
molecule. Considering protonation states 
(how the molecule changes when one proton is added
to it) and tautomer states (how the molecule can change when a proton
migrates from one part of the molecule to another), the challenge
increases. The need for a unique representation that is invariant with
respect to the atom ordering arises due to the fact that graph
isomorphism (i.e., checking whether two structures represent the same molecule) is an
expensive operation. The first such ``unique representation
algorithm,'' or \emph{canonicalization} algorithm, was described by
Morgan \cite{Morgan1965}, allowing one to generate unique string
representations of chemical graphs letting one compare structures via
string comparisons. The SMILES (Simplified Molecular Input Line
Specification) format as defined by Weininger in 1988 
\cite{Weininger:1988kx} is an example of a
representation that can be canonicalized. Since the original
canonicalization algorithm was proprietary, multiple implementations
of the format have become available, each one employing a different
canonicalization algorithm, usually based on the Morgan algorithm. See
Warr~\cite{Warr:2011vn} for an extensive discussion on chemical
structure representations.

@@

Graph substructure matching is a variant of \emph{graph isomorphism},
which is widely believed to be computationally
intractable~\cite{cordella2001}. To execute a graph isomorphism search
across a full chemical database of thousands or even millions of
structures is infeasible \cite{Weininger:2011ly}. Speedups can be
obtained via the use of heuristics such as structural
\emph{fingerprints}. Fingerprints encode characteristic features of a
given chemical structure, usually in a fixed-length
bitmap. Fingerprints fall broadly into two categories: structure keys
and hashed keys. In structure keys, each bit position corresponds to a
distinct substructure such as a functional group. Examples include MACCS and PubChem
keys. Hashed keys are those in which substructural patterns
are represented as strings and then hashed to a random bit
position. As a result, a given position can encode multiple
substructures. The advantage of such fingerprints is that they can
cover an arbitrarily large collection of substructures such as ``paths
of length $N$,'' or circular environments. Examples include
ECFPs (Extended Connectivity Fingerprints, which use local topological
information) and Daylight fingerprints (which are ``folded'' to
optimize information density and screening speed).
Given a binary fingerprint, we can first ``pre-screen'' a database, to
ignore molecules that cannot possibly match the query,
 by requiring that all bits in a query fingerprint must also
be present in the target fingerprint. Since the target fingerprints
are pre-computed, this check can be performed extremely
rapidly on modern hardware. As a result, we apply the actual
isomorphism test only on those molecules that pass the
screen. Fingerprints can also be used to rapidly search databases for
similar molecules, by using a similarity metric such as the
Tanimoto coefficient to compare the query and target
fingerprints. Additional heuristics have been developed
\cite{Swamidass:2007ve} to further speed up similarity searches.

@@

\bibliographystyle{abbrv}
\bibliography{resubmission}

>>>>

\documentclass{sig-alternate}

@@

\usepackage[numbers,sort&compress]{natbib}

@@

So why is chemical information important and why should we care about
its public availability -- and how does this relate to the field of
computer science? Cheminformatics is widely used in fields
from agrochemical research to the design of
novel materials -- but we have chosen to use drug discovery as the context
of this article due to its central relevance for human well-being.
The art and science of drug discovery centers around small
molecules, with a core component of methods involving
techniques and algorithms to handle, analyse and interpret
chemical structure information.
Unfortunately, drug discovery research is slow, expensive and prone to failure.
Public availability of chemical information and
tools to handle it are important, as the more information available to each
researcher, the better the chances of avoiding the many causes of attrition.
Computer science research is highly relevant to managing the \textit{volume}
and \textit{complexity} of chemical information.
A single database such as PubChem contains more than 34 million
chemical structure records), along with an even larger number of
\emph{annotations}, such as synonyms (different names for the same molecule;
PubChem stores 50.1 million synonyms for 19.6 million compounds),
known targets (drugs with which the molecule is known to interact),
mode of action (how the molecule interacts with its targets),
regulatory approval status and much more.

@@

Chemical and bioactivity
databases such as ChEMBLdb and PubChem are freely
available and contain in the order of millions (ChEMBLdb) and tens of
millions (PubChem) of data points. Integration of these disparate
data is essential for researchers to obtain the fullest possible
perspective on what is presently known, tracking ongoing advances in
science as these become available. Integration of data
across chemical databases is a challenge due to sheer data volume and
due to the difficulties in normalization of chemical and bioactivity
data. Molecular graphs must be captured
in a machine-readable fashion. It is also necessary to be able to
search in these data, for example multi-labeled graphs, where graph labels can change
from database to database. Not only can labels differ between databases,
but a chemical graph can be encoded in multiple ways, depending on how
the nodes are ordered, resulting in multiple representations of the same
molecule. Considering protonation states
(how the molecule changes when one proton is added
to it) and tautomer states (how the molecule can change when a proton
migrates from one part of the molecule to another), the challenge
increases. The need for a unique representation that is invariant with
respect to the atom ordering arises due to the fact that graph
isomorphism (i.e., checking whether two structures represent the same molecule) is an
expensive operation. The first such ``unique representation
algorithm,'' or \emph{canonicalization} algorithm, was described by
Morgan \cite{Morgan1965}, allowing one to generate unique string
representations of chemical graphs letting one compare structures via
string comparisons. The SMILES (Simplified Molecular-Input Line-Entry
System) format as defined by Weininger in 1988
\cite{Weininger:1988kx} is an example of a
representation that can be canonicalized. Since the original
canonicalization algorithm was proprietary, multiple implementations
of the format have become available, each one employing a different
canonicalization algorithm, usually based on the Morgan algorithm. See
Warr~\cite{Warr:2011vn} for an extensive discussion on chemical
structure representations.

@@

Graph substructure matching is a variant of \emph{graph isomorphism},
which is widely believed to be computationally
intractable~\cite{cordella2001}. To execute a graph isomorphism search
across a full chemical database of thousands or even millions of
structures is unfeasible \cite{Weininger:2011ly}. Speedups can be
obtained via the use of heuristics such as structural
\emph{fingerprints}. Fingerprints encode characteristic features of a
given chemical structure, usually in a fixed-length
bitmap. Fingerprints fall broadly into two categories: structure keys
and hashed keys. In structure keys, each bit position corresponds to a
distinct substructure such as a functional group. Examples include MACCS and PubChem
keys. Hashed keys are those in which substructural patterns
are represented as strings and then hashed to a random bit
position. As a result, a given position can encode multiple
substructures. The advantage of such fingerprints is that they can
cover an arbitrarily large collection of substructures such as ``paths
of length $N$,'' or circular environments. Examples include Daylight fingerprints (which are ``folded'' to
optimize information density and screening speed) and
ECFPs (Extended Connectivity Fingerprints, which use local topological
information).
Given a binary fingerprint, we can first ``pre-screen'' a database, to
ignore molecules that cannot possibly match the query,
 by requiring that all bits in a query fingerprint must also
be present in the target fingerprint. Since the target fingerprints
are pre-computed, this check can be performed extremely
rapidly on modern hardware. As a result, we apply the actual
isomorphism test only on those molecules that pass the
screen. Fingerprints can also be used to rapidly search databases for
similar molecules, by using a similarity metric such as the
Tanimoto coefficient to compare the query and target
fingerprints. Additional heuristics have been developed
\cite{Swamidass:2007ve} to further speed up similarity searches.

@@

\bibliographystyle{abbrv}
\bibliography{CACM_second_revision}
